---
title: "part b draft xxxx"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
date: "2025-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Loading the Required Packages
```{r}
pacman::p_load(tidyverse, dplyr, readr, here, sf, tmap, corrplot, spdep, janitor, here, pheatmap, caret, car)
```


```{r}
library(MASS)
```
```{r}
library(dplyr) 
```


# Loading the Datasets

* This script loads the Tesco grocery data, MSOA demographic data, and the London shapefile, cleans all their MSOA names and codes, and matches them so every dataset lines up properly. 
* It removes unwanted rows, checks which MSOAs matched or didnt, and joins everything into one complete dataset that includes map shapes, food purchases, and demographic indicators. After that, it drops the geometry to create a clean modelling table, removes missing values, and saves both the spatial file and the merged CSV. 
* Finally, the script loops through each predictor variable, creates a scatterplot with a fitted regression line against energy density, and saves every plot as a PNG in the figs folder.

```{r}


# 0) Libraries
library(readr)
library(dplyr)
library(stringr)
library(janitor)
library(sf)
library(ggplot2)

# 1) File paths (edit if folders differ)
path_tesco <- "data/year_msoa_grocery.csv"
path_atlas <- "data/msoa-data.csv"   # CSV downloaded from MSOA Atlas
path_shp   <- "data/statistical-gis-boundaries-london/statistical-gis-boundaries-london/ESRI/MSOA_2011_London_gen_MHW.shp"

# 2) Tesco (has MSOA code)
tesco <- readr::read_csv(path_tesco, show_col_types = FALSE) |>
  dplyr::rename(msoa_code = area_id) |>
  dplyr::mutate(msoa_code = toupper(stringr::str_trim(as.character(msoa_code))))

# 3) Atlas (handle encoding; clean names)
atlas <- readr::read_csv(path_atlas, locale = readr::locale(encoding = "UTF-8"), show_col_types = FALSE) |>
  janitor::clean_names()


if (!"msoa_name" %in% names(atlas)) stop("Atlas file must contain a column named 'msoa_name'.")

# Remove aggregate rows such as “AVERAGE”, “ALL LONDON”, etc.
atlas <- atlas |>
  dplyr::filter(!stringr::str_detect(stringr::str_to_upper(coalesce(msoa_name, "")),
                                     "AVERAGE|TOTAL|ALL LONDON|GREATER LONDON"))

# 4) Shapefile (2011). Normalise to msoa_code/msoa_name
msoas_raw <- sf::st_read(path_shp, quiet = TRUE)

# Try common field names (works for 2011/2021 shapefiles too if you swap files later)
code_col <- intersect(names(msoas_raw), c("msoa_code", "MSOA21CD", "MSOA11CD"))[1]
name_col <- intersect(names(msoas_raw), c("msoa_name", "MSOA21NM", "MSOA11NM"))[1]
if (is.na(code_col) || is.na(name_col)) {
  stop(sprintf("Could not find MSOA code/name fields in the shapefile. Found: %s",
               paste(names(msoas_raw), collapse = ", ")))
}

msoas <- msoas_raw |>
  dplyr::rename(msoa_code = !!code_col, msoa_name = !!name_col) |>
  dplyr::mutate(msoa_code = toupper(stringr::str_trim(as.character(msoa_code))))

# 5) Consistent uppercase names (lookup)
msoas_lookup <- msoas |>
  sf::st_drop_geometry() |>
  dplyr::mutate(
    msoa_name = msoa_name |>
      stringr::str_to_upper() |>
      stringr::str_replace_all("&", "AND") |>
      stringr::str_replace_all("[^A-Z0-9 ]", " ") |>
      stringr::str_squish()
  ) |>
  dplyr::select(msoa_name, msoa_code)

atlas_for_join <- atlas |>
  dplyr::mutate(
    msoa_name = msoa_name |>
      stringr::str_to_upper() |>
      stringr::str_replace_all("&", "AND") |>
      stringr::str_replace_all("[^A-Z0-9 ]", " ") |>
      stringr::str_squish()
  )

# 6) Join diagnostics 
cat("# rows — tesco:", nrow(tesco), " atlas:", nrow(atlas), " msoas:", nrow(msoas), "\n")

atlas_with_code <- dplyr::inner_join(atlas_for_join, msoas_lookup, by = "msoa_name")
cat("Matched atlas→MSOA by name:", nrow(atlas_with_code), "rows\n")

unmatched_atlas <- dplyr::anti_join(atlas_for_join, msoas_lookup, by = "msoa_name")
cat("Unmatched atlas names:", nrow(unmatched_atlas), "\n")
if (nrow(unmatched_atlas) > 0) readr::write_csv(unmatched_atlas, "data/_unmatched_atlas_names.csv")

# 7) Tesco + Atlas by code
tesco_atlas <- dplyr::inner_join(
  tesco,
  dplyr::select(atlas_with_code, msoa_code, dplyr::everything()),
  by = "msoa_code"
)
cat("Rows after Tesco+Atlas join:", nrow(tesco_atlas), "\n")

# 8) Join with shapefile (sf)
msoa_all <- dplyr::inner_join(msoas, tesco_atlas, by = "msoa_code")
cat("Rows after spatial join (should be MSOA count):", nrow(msoa_all), "\n")

# 9) Plain data frame for modelling/plots
stats_data <- sf::st_drop_geometry(msoa_all)

# 10) Remove missing values from the stats_data
stats_data_clean <- na.omit(stats_data)  # Remove rows with any missing values

# Save clean outputs
readr::write_csv(stats_data, "data/merged_tesco_atlas.csv")
sf::st_write(msoa_all, "data/msoa_all.gpkg", delete_dsn = TRUE, quiet = TRUE)

# 11) Scatterplot loop
predictors <- c(
  "avg_age","people_per_sq_km","age_65+","man_day","transaction_days",
  "num_transactions","population","male","female",
  "f_readymade","f_wine","f_soft_drinks","f_fruit_veg","f_sweets",
  "f_meat_red","f_dairy","f_eggs","f_fish","f_grains","f_poultry","f_water"
)
predictors <- intersect(predictors, names(stats_data))

if (!dir.exists("figs")) dir.create("figs", recursive = TRUE)

coerce_num <- function(v) { if (is.numeric(v)) v else suppressWarnings(as.numeric(v)) }

for (x in predictors) {
  tmp <- stats_data[, c(x, "energy_density")]
  tmp[[x]] <- coerce_num(tmp[[x]])
  tmp[["energy_density"]] <- coerce_num(tmp[["energy_density"]])

  p <- ggplot2::ggplot(tmp, ggplot2::aes(x = .data[[x]], y = .data[["energy_density"]])) +
    ggplot2::geom_point(alpha = 0.6, na.rm = TRUE) +
    ggplot2::geom_smooth(method = "lm", se = TRUE) +
    ggplot2::labs(
      title = paste("Energy density vs", x),
      x = x,
      y = "Energy density (kcal/g)"
    ) +
    ggplot2::theme_minimal()

  print(p)
  ggplot2::ggsave(
    file.path("figs", paste0("scatter_", gsub("[^A-Za-z0-9_]", "_", x), ".png")),
    p, width = 7, height = 5, dpi = 300
  )
}

```



# Map Of Engergy Density before modeling.

```{r}
# ---- ENERGY DENSITY MAP (Before Modelling) ----
library(tmap)
library(dplyr)

# Ensuring correct column name for code
names(msoa_all)[grepl("msoa_code", names(msoa_all), ignore.case = TRUE)] <- "msoa_code"

# Checking that energy_density exists
if(!"energy_density" %in% names(msoa_all)) {
  stop("energy_density column not found in msoa_all dataset.")
}

# Plot map
tmap_mode("plot")

energy_map <- tm_shape(msoa_all) +
  tm_polygons("energy_density",
              style = "quantile",
              n = 5,
              palette = "YlOrRd",
              title = "Energy Density (kcal/g)") +
  tm_layout(main.title = "Regional Distribution of Energy Density (Pre-Model)",
            legend.outside = TRUE)

# Displaying map
energy_map

# Saving high-quality PNG
if (!dir.exists("Figures_final")) dir.create("Figures_final", recursive = TRUE)
tmap::tmap_save(energy_map, "Figures_final/energy_density_pre_model.png",
                width = 2200, height = 1600, dpi = 220)

```


# Checking the Dimensions of Dataset

```{r}

dim(stats_data)        # rows, columns
ncol(stats_data)       # just the number of variables
names(stats_data)[1:20] # peek at the first 20 column names
```

# Applying the Stepwise Regression to the Varibales

```{r}
# ===============================
# Stepwise Regression on  Variables
# ===============================

library(MASS)  # For stepAIC

# 0) Safety checks 
stopifnot(exists("stats_data"))
stopifnot("energy_density" %in% names(stats_data))

# 1) List of variables
selected_variables <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children",
  "household_composition_2011_percentages_lone_parent_household",
  "household_composition_2011_percentages_one_person_household",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language",
  "qualifications_2011_census_no_qualifications",
  "qualifications_2011_census_schoolchildren_and_full_time_students_age_18_and_over",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12",
  "obesity_percentage_of_the_population_aged_16_with_a_bmi_of_30_modelled_estimate_2006_2008",
  "people_per_sq_km", 
  "male"
  
)

# 2) Keeping only the variables 
model_data <- stats_data[, c("energy_density", selected_variables), drop = FALSE]

# 3) Cleaning data (remove NAs for the selected variables)
model_data_clean <- na.omit(model_data)

# 4) Runing stepwise regression
# Useing both directions: forward and backward selection
stepwise_model <- stepAIC(lm(energy_density ~ ., data = model_data_clean), direction = "both", trace = FALSE)

# 5) Getting the selected variables after stepwise regression
selected_vars_stepwise <- attr(terms(stepwise_model), "term.labels")
cat("Final selected variables after stepwise regression:\n")
print(selected_vars_stepwise)

# 6) Getting the model summary
summary(stepwise_model)

# 7) Saving the results
if (!dir.exists("tables")) dir.create("tables", recursive = TRUE)
if (!dir.exists("figs"))   dir.create("figs", recursive = TRUE)

# Saving the coefficients of the final model
coef_table <- broom::tidy(stepwise_model)
readr::write_csv(coef_table, "tables/stepwise_model_coefficients.csv")

# Plot the coefficients of the selected variables (final model)
p_coef <- ggplot(coef_table, aes(x = reorder(term, estimate), y = estimate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Stepwise Model: Standardised Coefficients",
       x = "Variables", y = "Standardised Coefficient") +
  theme_minimal(base_size = 12)

# Save the plot
ggsave("figs/stepwise_coefficients.png", p_coef, width = 8, height = 6, dpi = 300)

# Print final selected variables and model summary
cat("\nSelected variables:\n")
print(selected_vars_stepwise)

cat("\nModel Summary:\n")
summary(stepwise_model)

```

# Interpretation of the Stepwise Regression Output

The regression output provides a comprehensive overview of the relationship between the selected variables and energy density (in kcal/g). Lets break down the key results:

Model Summary:

Multiple R-squared: 0.4455

This means that 44.55% of the variance in energy density is explained by the selected predictors. This is a decent explanatory power, but there’s room for improvement.

Adjusted R-squared: 0.4405

This metric adjusts for the number of predictors and confirms that the model remains robust with these variables.

F-statistic: 89.49 (p-value < 2.2e-16)

The F-test confirms that the model as a whole is statistically significant (p-value < 0.05), meaning that at least some of the predictors are significantly related to energy density.



# Getting the Coeffient Values from the Stepwise Regression

```{r}
# Getting the coefficients from the stepwise model
coef_table <- broom::tidy(stepwise_model)

# Removing the intercept and filter significant terms (p < 0.05)
coef_table_clean <- coef_table %>%
  filter(term != "(Intercept)" & p.value < 0.05) %>%
  arrange(estimate)

# Getting the confidence intervals for the coefficients
conf_intervals <- confint(stepwise_model)[-1, ]  # Remove the intercept row

# Ensuring that the number of rows in the confidence intervals matches the number of coefficients
if (nrow(coef_table_clean) != nrow(conf_intervals)) {
  stop("Mismatch in the number of rows between coefficients and confidence intervals.")
}

# Merging the coefficients with their confidence intervals
coef_table_clean$conf.low <- conf_intervals[, 1]
coef_table_clean$conf.high <- conf_intervals[, 2]

# Checking the final table to ensure conf.low and conf.high are added correctly
head(coef_table_clean)


```

# Creating the Coeffient Plot For the variables selected by Stepwise regression

```{r}
# Creating a mapping of long variable names to shorter labels
labels_map <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children" = "Couple household w/o dependents",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children" = "No adults in work (w/ dependents)",
  "household_composition_2011_percentages_one_person_household" = "One-person households (%)",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd" = "Median annual household income",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households" = "60+ living in pension credit households",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12" = "Obese children (Year 6)",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language" = "No English as main language",
  "household_composition_2011_percentages_lone_parent_household" = "Lone parent households (%)"
)

# Appling this mapping to the coefficient data
coef_table_clean$label <- labels_map[coef_table_clean$term]

# Displaying coefficients with more precision (up to 6 decimal places)
coef_table_clean$beta_lbl <- sprintf("%.6f", coef_table_clean$estimate)

# Creating the plot with standard ggplot 
ggplot(coef_table_clean, aes(x = reorder(label, estimate), y = estimate, 
                             ymin = conf.low, ymax = conf.high, fill = estimate > 0)) +
  geom_bar(stat = "identity") +
  geom_errorbar(width = 0.2) +
  # Adding coefficient values beside bars
  geom_text(aes(label = beta_lbl), 
            hjust = ifelse(coef_table_clean$estimate > 0, 1.1, -0.1), 
            color = "black", size = 3.5) +  # Positioning labels to avoid overlap
  coord_flip() +
  labs(title = "Stepwise Regression: Significant Coefficients",
       subtitle = "Standardised Coefficients with 95% Confidence Intervals",
       x = "Predictors", y = "Standardised Coefficient (β)") +
  scale_fill_manual(values = c("steelblue", "tomato")) + 
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Saving the plot to the 'figs' folder
ggsave("figs/stepwise_coefficients_with_values.png", width = 8, height = 6, dpi = 300)

```

# Creating the Scatter Plot for the selected variables

```{r}
# Final selected variables after stepwise regression
final_selected_variables <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children",
  "household_composition_2011_percentages_lone_parent_household",
  "household_composition_2011_percentages_one_person_household",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12"
)

# Ensuring the 'figs' folder exists
if (!dir.exists("figs")) dir.create("figs", recursive = TRUE)

# Looping through the selected variables and create individual scatter plots
for (var in final_selected_variables) {
  # Create the scatter plot for each variable against energy_density
  p <- ggplot(stats_data, aes_string(x = var, y = "energy_density")) +
    geom_point(color = "blue", alpha = 0.6, na.rm = TRUE) +  # Blue dots
    geom_smooth(method = "lm", se = TRUE, color = "red") +    # Red regression line
    labs(title = paste("Energy Density vs", var),
         x = var, y = "Energy Density (kcal/g)") +
    theme_minimal(base_size = 12)

  # Saving the plot in the 'figs' folder with a name corresponding to the variable
  ggsave(paste0("figs/scatter_", gsub("[^A-Za-z0-9]", "_", var), ".png"), p, width = 8, height = 6, dpi = 300)
  
  
  print(p)
}

```

# Creating Spatial Maps


```{r}

library(tmap)
library(sf)
library(dplyr)

# 1) Final selected vars 
final_selected_variables <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children",
  "household_composition_2011_percentages_lone_parent_household",
  "household_composition_2011_percentages_one_person_household",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12"
)

labels_map <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children" =
    "Couple hh w/o dependents (%)",
  "household_composition_2011_percentages_lone_parent_household" =
    "Lone-parent hh (%)",
  "household_composition_2011_percentages_one_person_household" =
    "One-person hh (%)",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language" =
    "No English main lang (%)",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children" =
    "No adults in work w/ dependents (%)",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd" =
    "Median hh income (£)",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households" =
    "60+ on pension credit (%)",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12" =
    "Year 6 obese (%)"
)

# 2) Ensuring we have an sf and numeric columns (for mapping)
stopifnot(inherits(msoa_all, "sf"))
msoa_all <- msoa_all |>
  mutate(across(all_of(final_selected_variables),
                ~ suppressWarnings(as.numeric(.))))


msoa_all <- msoa_all |>
  mutate(across(all_of(final_selected_variables),
                ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

# 3) Choosing a vivid palette (exact, valid names) and generate a color vector
if (!requireNamespace("cols4all", quietly = TRUE)) install.packages("cols4all")
library(cols4all)
pal_vals <- c4a("plasma", 5)          # <-- change to "viridis", "magma", "carto.tropic", etc. if you prefer

# 4) Map builder (tmap v4.2)
tmap_mode("plot")
if (!dir.exists("figs/maps")) dir.create("figs/maps", recursive = TRUE)

make_map <- function(var) {
  tm_shape(msoa_all) +
    tm_polygons(
      fill = var,
      fill.scale = tm_scale_intervals(
        style   = "quantile",   # 5 equal-frequency bins
        n       = 5,
        values  = pal_vals,     # pass the actual vector of colors
        value.na = "grey90"
      ),
      fill.legend = tm_legend(
        title = labels_map[[var]],
        frame = TRUE
      )
    ) +
    tm_borders(col = "white", lwd = 0.4) +
    tm_title(paste0(labels_map[[var]], " in London")) +
    tm_layout(legend.outside = TRUE, frame = FALSE)
}

# 5) Render + save all maps
for (var in final_selected_variables) {
  if (!var %in% names(msoa_all)) next
  p <- make_map(var)
  print(p)
  tmap_save(
    p,
    filename = paste0("figs/maps/", gsub("[^A-Za-z0-9]+", "_", var), "_map.png"),
    width = 2200, height = 1600, dpi = 220
  )
}

```


# Getting Names of a regions where each varibale is  High /Low 

```{r}

library(dplyr)
library(readr)
library(sf)

stopifnot(inherits(msoa_all, "sf"))

# Making sure these are numeric for ranking
msoa_all <- msoa_all %>%
  dplyr::mutate(dplyr::across(all_of(final_selected_variables), ~ suppressWarnings(as.numeric(.))))


msoa_name_col <- intersect(names(msoa_all), c("msoa_name","msoa_name.x","MSOA11NM","MSOA21NM"))[1]
lad_name_col  <- intersect(names(msoa_all), c("LAD11NM","LAD21NM","LAD23NM","lad_name","Borough","LA_NAME"))[1]

if (!dir.exists("tables")) dir.create("tables", recursive = TRUE)

for (var in final_selected_variables) {
  if (!var %in% names(msoa_all)) {
    message("Skipping (missing in data): ", var)
    next
  }

  # Building a safe selection vector for reporting columns
  cols_keep <- c("msoa_code", msoa_name_col, lad_name_col, var)
  cols_keep <- cols_keep[!is.na(cols_keep) & cols_keep %in% names(msoa_all)]

  df <- msoa_all %>%
    sf::st_drop_geometry() %>%
    dplyr::select(dplyr::all_of(cols_keep)) %>%
    # Give consistent column names for pretty output
    dplyr::rename(
      msoa      = dplyr::all_of(msoa_name_col),
      borough   = dplyr::all_of(lad_name_col)
    ) %>%
    dplyr::mutate(value = .data[[var]])

  # Top 10
  top_regions <- df %>%
    dplyr::arrange(dplyr::desc(.data$value)) %>%
    dplyr::slice_head(n = 10)

  # Bottom 10
  bottom_regions <- df %>%
    dplyr::arrange(.data$value) %>%
    dplyr::slice_head(n = 10)

  # Console summary
  cat("\n==========================================\n")
  cat("Variable:", labels_map[[var]] %||% var, "\n")
  cat("Top 10 regions (highest values):\n"); print(top_regions)
  cat("\nBottom 10 regions (lowest values):\n"); print(bottom_regions)
  cat("==========================================\n")

  # Save CSVs
  safe <- gsub("[^A-Za-z0-9]+", "_", var)
  readr::write_csv(top_regions,    file.path("tables", paste0("top10_", safe, ".csv")))
  readr::write_csv(bottom_regions, file.path("tables", paste0("bottom10_", safe, ".csv")))
}

```

# Creating side-by-side maps.

* This script creates two types of maps for each selected variable in the dataset a full map showing the value of the variable across all London MSOAs, and a high only map that highlights only the top 10 percent of areas while keeping the rest of London in light grey. 
* To do this, the code first converts all selected variables to numeric, defines a consistent colour palette and breakpoints, and calculates the 90th-percentile threshold for what counts as high.It then generates each full map using quantile-based colour breaks and an identical colour scale.
* For the high-only version, the script keeps the original values only for MSOAs above the threshold and blanks out the rest so only the highest areas are coloured. Finally, for every variable, the code arranges the full map and the high-only map side-by-side, displays them, and saves each combined figure as a PNG file in the figs/maps folder.

```{r}
library(tmap)
library(sf)
library(dplyr)
library(cols4all)


final_selected_variables <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children",
  "household_composition_2011_percentages_lone_parent_household",
  "household_composition_2011_percentages_one_person_household",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12"
)

labels_map <- c(
  "household_composition_2011_percentages_couple_household_without_dependent_children" = "Couple hh w/o dependents (%)",
  "household_composition_2011_percentages_lone_parent_household" = "Lone-parent hh (%)",
  "household_composition_2011_percentages_one_person_household" = "One-person hh (%)",
  "household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language" = "No English main lang (%)",
  "adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children" = "No adults in work w/ dependents (%)",
  "household_income_estimates_2011_12_total_median_annual_household_income_u_fffd" = "Median hh income (£)",
  "income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households" = "60+ on pension credit (%)",
  "obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12" = "Year 6 obese (%)"
)

# making sure numeric
stopifnot(inherits(msoa_all, "sf"))
msoa_all <- msoa_all %>%
  mutate(across(all_of(final_selected_variables), ~ suppressWarnings(as.numeric(.))))

# same palette as main maps
pal_vals <- c4a("plasma", 5)

# top 10% threshold
HIGH_QUANTILE <- 0.90

tmap_mode("plot")
if (!dir.exists("figs/maps")) dir.create("figs/maps", TRUE)


make_full_map <- function(var) {
  # breaks based on FULL data range
  brks <- quantile(msoa_all[[var]], probs = seq(0, 1, length.out = 6), na.rm = TRUE)

  tm_shape(msoa_all) +
    tm_polygons(
      fill = var,
      fill.scale = tm_scale_intervals(
        style   = "fixed",       # we now fix breaks explicitly
        breaks  = brks,
        values  = pal_vals,
        value.na = "grey90"
      ),
      fill.legend = tm_legend(
        title = labels_map[[var]],
        frame = TRUE
      )
    ) +
    tm_borders(col = "white", lwd = 0.4) +
    tm_title(paste0(labels_map[[var]], " in London")) +
    tm_layout(legend.outside = TRUE, frame = FALSE)
}

# -----------------------------
# 2) High-only map using SAME scale
# -----------------------------
make_high_only_map <- function(var) {
  stopifnot(var %in% names(msoa_all))

  # same global breaks as full map
  brks <- quantile(msoa_all[[var]], probs = seq(0, 1, length.out = 6), na.rm = TRUE)

  # threshold for "high" (top 10%)
  thr <- quantile(msoa_all[[var]], probs = HIGH_QUANTILE, na.rm = TRUE)

  high_col <- paste0(var, "__high_only")

  # keep original values for high areas, NA elsewhere
  msoa_plot <- msoa_all %>%
    mutate(!!high_col := ifelse(!is.na(.data[[var]]) & .data[[var]] >= thr,
                                .data[[var]], NA_real_))

  tm_shape(msoa_all) +
    # base layer: all MSOAs in light grey
    tm_polygons(col = "grey90", lwd = 0.2) +
    # overlay: only high areas coloured, with SAME breaks + colours
    tm_shape(msoa_plot) +
    tm_polygons(
      fill = high_col,
      fill.scale = tm_scale_intervals(
        style   = "fixed",
        breaks  = brks,
        values  = pal_vals,
        value.na = NA        # non-high = transparent
      ),
      fill.legend = tm_legend(
        title = paste0(labels_map[[var]], " (≥ ", round(thr, 2), ")"),
        frame = TRUE
      ),
      col = NA
    ) +
    tm_borders(col = "white", lwd = 0.3) +
    tm_title(paste0("High areas: ", labels_map[[var]],
                    " (Top ", round((1 - HIGH_QUANTILE) * 100), "%)")) +
    tm_layout(legend.outside = TRUE, frame = FALSE)
}

# -----------------------------
# 3) Side-by-side maps + save
# -----------------------------
for (var in final_selected_variables) {
  full_map <- make_full_map(var)
  high_map <- make_high_only_map(var)

  # view in R
  print(tmap_arrange(full_map, high_map, ncol = 2))

  # save combined figure
  safe <- gsub("[^A-Za-z0-9]+", "_", var)
  tmap_save(
    tmap_arrange(full_map, high_map, ncol = 2),
    filename = paste0("figs/maps/combined_", safe, ".png"),
    width = 3000, height = 1600, dpi = 220
  )
}

```


# Checking the Model performance by Training and Testing Sets

```{r}
set.seed(123)  # for reproducibility

# Split data
n <- nrow(model_data_clean)
train_idx <- sample(seq_len(n), size = 0.7 * n)

train_data <- model_data_clean[train_idx, ]
test_data  <- model_data_clean[-train_idx, ]

# Fitting model on training data
fit <- lm(formula(stepwise_model), data = train_data)

# Predicting on test data
test_data$predicted <- predict(fit, newdata = test_data)

# Calculating performance metrics
library(Metrics)
rmse_val <- rmse(test_data$energy_density, test_data$predicted)
mae_val  <- mae(test_data$energy_density, test_data$predicted)
r2_val   <- cor(test_data$energy_density, test_data$predicted)^2

cat("Test RMSE:", rmse_val, "\n")
cat("Test MAE:", mae_val, "\n")
cat("Test R² :", r2_val, "\n")

```

* The final stepwise regression model performed well, with an adjusted R squre of 0.4405.
This means that about 44% of the differences in energy density between regions can be explained by the sociodemographic factors included in the model.
In other words, people living conditions — such as their income, household type, and local deprivation levels — help us understand why some areas tend to buy more calorie-dense foods than others.

* When the data was divided into training 70% and testing 30% sets, the model continued to perform well.
It achieved a Test R squre of 0.453, meaning it could explain about 45% of the variation in new, unseen data — almost the same as the training result.
This consistency shows that the model is stable and not overfitted, which means it works well beyond the data it was trained on.

* The model prediction errors were small:

RMSE 0.057 and MAE 0.045 indicate that, on average, the model’s predicted energy density values were very close to the actual observed values.

Overall, this suggests that variables such as household composition types of families, income levels, deprivation, language background, and obesity rates are strong indicators of regional differences in dietary health risks.
Areas with lower incomes, higher deprivation, or more households without English as a main language tend to have higher average energy density, meaning their diets are generally less healthy.


# visualising the Train and test sets

```{r}
library(ggplot2)
p1 <- ggplot(test_data, aes(x = energy_density, y = predicted)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs Predicted Energy Density (Test Set)",
       x = "Actual Energy Density",
       y = "Predicted Energy Density") +
  theme_minimal(base_size = 12)
ggsave("figs/actual_vs_predicted_test.png", p1, width = 8, height = 6, dpi = 300)
print(p1)

```

* This scatter plot shows how well the model’s predicted energy density values match the actual values from the test data.
Each blue dot represents one region in London.

* The x-axis shows the actual energy density observed in the data.

* The y-axis shows the predicted energy density from the model.

* The red dashed line represents perfect predictions if the model were 100% accurate, all points would fall exactly on this line.

* In this plot, most points are close to the red line, which means the model’s predictions are quite accurate.
There is some natural scatter, which is expected with real-world social data, but there are no extreme outliers.
This shows that the model predicts energy density values consistently and without major bias (it doesn’t systematically overpredict or underpredict).


# Residual Plot

```{r}
p2 <- ggplot(test_data, aes(x = predicted, y = energy_density - predicted)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Predicted (Test Set)",
       x = "Predicted Energy Density",
       y = "Residuals") +
  theme_minimal(base_size = 12)
ggsave("figs/residuals_vs_predicted.png", p2, width = 8, height = 6, dpi = 300)
print(p2)

```

* This plot shows the residuals (the difference between actual and predicted energy density) against the predicted values from the model.

* The horizontal red dashed line represents zero error where the model’s predictions are exactly correct.

* The orange points represent the prediction errors for each region.

* Most of the points are evenly scattered around the red line, with no clear upward or downward pattern.
* This means the model’s errors are randomly distributed, which is a good sign.
It shows that the model does not systematically overestimate or underestimate energy density in certain ranges.

* There are a few points slightly above or below the line, which is expected in real data, but overall the spread is narrow.
This indicates that the model fits the data consistently and without major bias across all predicted values.

# Comparing the Training and testing sets.

```{r}
set.seed(123)

# Split data 70–30
n <- nrow(model_data_clean)
train_idx <- sample(seq_len(n), size = 0.7 * n)

train_data <- model_data_clean[train_idx, ]
test_data  <- model_data_clean[-train_idx, ]

# Fit model on training data
fit <- lm(formula(stepwise_model), data = train_data)

# Predict on both training and test data
train_data$predicted <- predict(fit, newdata = train_data)
test_data$predicted  <- predict(fit, newdata = test_data)

# Load Metrics package
library(Metrics)

# Compute metrics for both sets
train_rmse <- rmse(train_data$energy_density, train_data$predicted)
train_mae  <- mae(train_data$energy_density, train_data$predicted)
train_r2   <- cor(train_data$energy_density, train_data$predicted)^2

test_rmse <- rmse(test_data$energy_density, test_data$predicted)
test_mae  <- mae(test_data$energy_density, test_data$predicted)
test_r2   <- cor(test_data$energy_density, test_data$predicted)^2

# Combine metrics for comparison
comparison <- data.frame(
  Dataset = c("Training", "Testing"),
  RMSE = c(train_rmse, test_rmse),
  MAE = c(train_mae, test_mae),
  R2 = c(train_r2, test_r2)
)

print(comparison)

# ----------------------------------------------
# Visual comparison: Actual vs Predicted plots
# ----------------------------------------------
library(ggplot2)

ggplot() +
  geom_point(data = train_data, aes(x = energy_density, y = predicted), 
             color = "blue", alpha = 0.5, size = 2) +
  geom_point(data = test_data, aes(x = energy_density, y = predicted), 
             color = "red", alpha = 0.5, size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  labs(
    title = "Actual vs Predicted Energy Density (Training vs Test)",
    x = "Actual Energy Density",
    y = "Predicted Energy Density",
    caption = "Blue = Training | Red = Test"
  ) +
  theme_minimal(base_size = 13)

ggsave(
  filename = "Final Figures/actual_vs_predicted_train_test.png",
  width = 10, height = 7, dpi = 300
)
```

* Figure X compares the actual and predicted energy density values for both the training (blue) and test (red) datasets.
The clustering of points around the 1:1 dashed line indicates that the model predicts energy density accurately across regions.

* The close overlap between training and testing observations demonstrates consistent predictive behaviour across datasets, confirming that the stepwise regression model generalises well and does not exhibit overfitting.

* Minor deviations from the diagonal represent normal residual variation, but no systematic bias is evident, supporting the model’s reliability and stabil



# Comparing the Actual Energy Density and the Predicted Energy Density

```{r}
# ================================================================
# Clear Actual vs Predicted Maps (no NAs, bright colours)
# ================================================================
library(dplyr)
library(sf)
library(tmap)
library(classInt)

# --- 1) Prepare modelling dataset and predictions ---
df_model <- stats_data %>%
  dplyr::select(
    msoa_code, energy_density,
    household_composition_2011_percentages_couple_household_without_dependent_children,
    household_composition_2011_percentages_lone_parent_household,
    household_composition_2011_percentages_one_person_household,
    household_language_2011_percent_of_households_where_no_people_in_household_have_english_as_a_main_language,
    adults_in_employment_2011_census_percent_of_households_with_no_adults_in_employment_with_dependent_children,
    household_income_estimates_2011_12_total_median_annual_household_income_u_fffd,
    income_deprivation_2010_percent_of_people_aged_over_60_who_live_in_pension_credit_households,
    obesity_percent_of_measured_children_in_year_6_who_were_classified_as_obese_2009_10_2011_12
  ) %>%
  tidyr::drop_na()   # remove any rows with missing values

# --- 2) Fit model (same formula as your stepwise one) ---
fit_clean <- lm(formula(stepwise_model), data = df_model)

# --- 3) Add predictions ---
df_map <- df_model %>%
  mutate(predicted = predict(fit_clean, newdata = df_model),
         actual = energy_density) %>%
  dplyr::select(msoa_code, actual, predicted)

# --- 4) Join to spatial data ---
map2 <- msoa_all %>%
  left_join(df_map, by = "msoa_code") %>%
  tidyr::drop_na(actual, predicted)   # ensures only complete areas are shown

# --- 5) Consistent colour breaks across both maps ---
vals <- c(map2$actual, map2$predicted)
brks <- classInt::classIntervals(vals, n = 5, style = "quantile")$brks

# Bright, publication-friendly colour palette (alternatives: "viridis", "inferno", "roma")
pal <- "batlow"

tmap_mode("plot")

map_actual <- tm_shape(map2) +
  tm_polygons(
    fill = "actual",
    fill.scale = tm_scale_intervals(
      breaks = brks, values = pal
    ),
    fill.legend = tm_legend(title = "Actual Energy Density (kcal/g)")
  ) +
  tm_borders("white", lwd = 0.4) +
  tm_title("Actual Energy Density (Observed)") +
  tm_layout(frame = FALSE, legend.outside = TRUE)

map_predicted <- tm_shape(map2) +
  tm_polygons(
    fill = "predicted",
    fill.scale = tm_scale_intervals(
      breaks = brks, values = pal
    ),
    fill.legend = tm_legend(title = "Predicted Energy Density (kcal/g)")
  ) +
  tm_borders("white", lwd = 0.4) +
  tm_title("Predicted Energy Density (Model Output)") +
  tm_layout(frame = FALSE, legend.outside = TRUE)

# --- 6) Display and save ---
panel <- tmap_arrange(map_actual, map_predicted, ncol = 2)
print(panel)

if (!dir.exists("figs/maps")) dir.create("figs/maps", recursive = TRUE)
tmap_save(panel, "figs/maps/Actual_vs_Predicted_EnergyDensity_clean.png",
          width = 3200, height = 1500, dpi = 300)

```

* To visualise how well the model captured regional variation in dietary energy density, both the observed (actual) and predicted values were mapped across London’s MSOA regions. The analysis used the merged dataset that combined Tesco grocery data (providing energy density) with MSOA-level sociodemographic indicators from the London Atlas. From this dataset, only the variables identified as significant in the stepwise regression were retained, and all rows with missing values were removed to ensure accuracy.

* Using the final regression model, predicted energy density values were generated for each MSOA region based on its demographic and socioeconomic characteristics. These predictions were then linked back to the corresponding geographic boundaries in the MSOA shapefile to enable spatial visualisation.

* Both the actual and predicted energy density maps were created using the tmap package in R. Identical colour scales and quantile based classification breaks were applied to both maps to ensure direct visual comparability. Areas with higher predicted or actual energy density values are shown in darker shades, highlighting regions where diets are likely to be more energy dense. By comparing the two maps side by side, it is possible to assess how closely the model’s predictions align with the real spatial distribution of dietary energy density across London.



